This is an example of article title

### 2023-07-01 paper/paper.tex
chore: Overview

\section{Overview}
To partition a cloud of points, \verb|AdaptiveHierarchicalRegularBinning.jl| maps each
point to a $1$ dimensional space filling curve, Morton or Z-order curve. The hierarchical
tree is defined by the Morton index of each point.

\begin{example}
  For a $2$-dimensional space a target spatial tree with $2$ levels (other than the root).
  Assume we compute a morton index of $(0100)_2$. The 2 most significant bits address the
  top most level of the tree (the one directly under root).

  The most significant bit
  addresses the first dimension and the value $(0)_2$ signifies that the point lies in
  the left-most half of the dimension.

  The second most significant bit addresses the
  second dimension and a value of $(1)_2$ signifies that the point lies in the right-most
  half of the dimension (typicaly the lower-most half since this is $2$-dimensional,
  right-most is a more generic term used for any number of dimensions).

@@

Sorting in terms of these morton indices group the points togather for efficient access
when the tree is iterated over. One natural way of sorting these indices is the
\verb|Radixsort| where we sort based on the most significant radix and recurse for the
remaining bits in the same fashion. This type of sorting process respects the nature of
the morton index and thus can offer benefits over other sorting algorithms.

One benefit of the \verb|Radixsort| is that it offers partial sorting capabilities. When
the number of items within a bin is not high enough, there is no reason to continue the
spatial partitioning and therefore there is no reason to continue sorting. Stopping
\verb|Radixsort| at a specific level of recursion translates to stoping the tree at that
level, there is a $1-1$ corespondance of the \verb|Radixsort| recursion and the tree.

\verb|Radixsort| can also be trivialy parallelized, after sorting the based on the most
significant radix the sub-arrays left to sort are mutually independant and thus offer
themselves to easy parallelization.

Sorting based on a radix is not trivial. Any sorting algorithm will work just fine but
since small radix means small number of discrete values \verb|Countsort| is used.
\verb|Countsort| offers trades-off spatial complexity for runtime complexity.
The worst case runtime complexity of the algorithm is $O(n)$, which means that
it scales linearly with the size of the array to sort. While the spatial complexity for
the in-place variant is $O(max(r))$ where $r$ is the selected radix and $max(r)$ is the
maximum value allowed by $r$.

\verb|Countsort| can be parallelized as well, though not trivialy. The parallel variant
requires a larger spatial complexity of $O(t\cdot max(r))$ where $t$ is the number of
tasks allocated for the parallelization. The runtime complexity drops by the same factor
$t$, $O(n/t)$. Spatial complexity is a limites the capabilities of the algorithm but even
for large values of $t$ this auxilary space needed is not a bad trade-off.

An other trade-off to note is the selection of radix. The bigger the radix the larger
the load on the \verb|Countsort|. Since the time complexity of \verb|Countsort| is $O(n)$
and even better for the parallel variant $O(n/t)$ we prefer that most of the work is
delegated to \verb|Countsort| instead of the recursion process of \verb|Radixsort|.
Selecting a large radix though, might have an adverse effect. The auxilary space needed
is based on the max value of the radix which exponatialy grows as the radix grows, meaning
that we quickly run out of memory for larger matrices.

Our implementation is radix-agnostic, meaning that the user can select the radix. But the
default approach is a radix of $16$ bits. $16$ bits means that at the worst case of
morton indices of $128$ bits we will need to recurse $8$ times and the auxilary space needed
is a factor of $2^{16} = 64kB$.

@@

>>>>

\section{Overview}
To partition a cloud of points of $d$ dimensions,
\verb|AdaptiveHierarchicalRegularBinning.jl| maps each point to a $1$ dimensional
space filling curve, Morton or Z-order curve. The hierarchical tree is defined by the
Morton index of each point.

\subsection{Morton encoding}
Morton indices are binary words grouped into $l$ levels of $d$ bits each. Each level
describes in an increasingly finer detail the area in which the point resides. The
first level, comprised of the $d$ most significant bits of the morton index, describes
only in which half of each dimension the point lies.

$$b = \left\{\begin{matrix}
  (0)_2, & \textrm{Left-most half}  \\
  (1)_2, & \textrm{Right-most half}
\end{matrix}\right.$$

\begin{definition}[Left/Right-mostness]
  For a $1$-dimensional simply connected space $A = [m, M)$, we say that
  $x$ lies in the Left-most half, if and only if, $x<h=\frac{M+m}{2}$.
  Otherwise $x$ lies in the Right-most half.

  $$
  \begin{matrix}
    L &\triangleq& [m, h) \\
    R &\triangleq& [h, M)
  \end{matrix}
  $$

  \begin{figure}
    \centerline{
      \includegraphics[width=10pc]{figures/spatial-division.png}
      \includegraphics[width=10pc]{figures/spatial-tree.png}
    }
    \caption{ (a) Spatial division, (b) Spatial tree }
    \label{spatial-tree}
  \end{figure}

\end{definition}

\begin{example}[Morton index interpretation]
  Figure \ref{spatial-tree} depicts a $2$-dimensional space and a target spatial
  tree of at most $2$ levels (other than the root). The morton index, in such a tree,
  will have a length of $2\cdot2=4$. Assume we compute a morton index of $(0011)_2$.
  The first group is comprised of the $2$ most significant bits and addresses the first
  level of the tree (the one directly under root).
  \\\\
  The most significant bit addresses the first dimension and the second most significant
  bit addresses the second dimension. The value $(0)_2$ signifies that the point lies in
  the left-most half of each dimension.
  \\\\
  The next group describes the same quantities only a level lower. Left/Right-mostness
  is defined in the context of that level. Both values are $(1)_2$ so the point lies in
  the right-most half of both dimensions in the second level.
  \\\\
  Summarizing, the point belongs to the:
  \begin{itemize}
    \item right-most half of the left-most half of the first dimension.
    \item right-most half of the left-most half of the second dimension.
  \end{itemize}

@@

\subsection{Reordering}
Sorting in terms of the morton indices groups the points together for efficient access
when the tree is iterated over. We use the \verb|MSD Radixsort| algorithm to perform the
soring.

\subsubsection{MSD Radixsort}
\verb|MSD Radixsort| sorts based on the most significant radix. Afterwards, elements
with the same most significant digit (MSD) form a consecutive sub-array. The algorithm
recurses for each such sub-array selecting the next digit in line. Selecting a large
radix is crucial for the algorithm since a large radix lowers the number of recursions.
\\\\
\verb|MSD Radixsort| respects the hierarchical nature of the target tree and thus it offers
benefits over other sorting algorithms. Each recursion of the algorithm sorts a section
of the resulting tree. If the radix matches the number of dimensions then each recursion
sorts the current level of the tree.
\\\\
When the number of elements of a sub array is small the recursion for that sub-array stops.
This leads to a partialy sorted array by design. Not sorting small sub-arrays leads to less
overhead without sacrificing future performance. Nodes with unsorted morton indices will
always be leaf nodes.
\\\\
\verb|MSD Radixsort| can be trivialy parallelized. After the initial sorting, the formed
sub-arrays are mutually exclusive, meaning that no data dependencies are shared between
them. Each sub-array can then be sorted in parallel.

\subsubsection{Countsort}
\verb|MSD Radixsort| needs a sorting algorithm to sort the selected radix in each recursion.
\verb|Countsort| is a great fit for this type of sorting. \verb|Countsort| trades
space complexity for runtime complexity. The algorithm uses an auxilary vector, whose
size scales linearly with the selected radix $O(r)$. The benefit is the runtime
complexity is order $O(n + r)$ where $n$ is the size of the array to sort. Typically, $n \gg r$
and so the runtime complexity reduces to $O(n)$.
\\\\
\verb|Countsort| can be parallelized as well, though not as trivialy. The parallel version
of \verb|Countsort| requires space complexity of $O(t\cdot r)$ and runtime complexity of
$O(n/t + t\cdot r)$ where $t$ is the number of threads in the system. As long as $n$ is
sufficiently large $n \gg t^2\cdot r$ then the parallel implementation is preferable since
the runtime complexity reduces to $O(n/t)$.

\subsubsection{Radix trade-off}
While \verb|MSD Radixsort| fairs better for larger radices, \verb|Countsort| reduces in
performance.

\begin{example}
  For a $128$-bit morton index, a radix of $16$-bits is selected.
  The number of recursions in this case is $8$ which is fairly small and the condition
  $n\gg t^2\cdot r$ can still be satisfied.
  % TODO: Graphs runtime complexity vs r
\end{example}

@@

\section{Experiments}

\subsection{Uniform data 2 dimensions}
\subsection{Uniform data 10 dimensions}
\subsection{MNIST data 10 dimensions}
\subsection{PBMC data 10 dimensions}
\subsection{PBMC 36 dimensions}
